---
title: "DataManagement"
author: "Wendy Leuenberger"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, include = FALSE}
library(magrittr)
library(tidyverse)
library(readxl)
library(lubridate)
```

```{r figure settings, include = FALSE}
tbw = theme_bw(base_size = 18)
th = theme(panel.grid = element_blank())
```


## Read in data

Note: This R file should be located in the same working directory as the data. Otherwise reading data from other places in the google drive is challenging and I couldn't get it to work. Might be tricky since the drive is "shared with me" not a shared drive or my drive.

```{r, warning = FALSE, message = FALSE}
Data2017 <- read_csv('test_form5_2019_12_19_14_56_01_542496.csv')
Data2018 <- read_csv(
  'Form_BS162_2018_v2_2019_12_19_14_56_54_642030.csv')
Data2019 <- read_csv(
  'Form_BS162_2019_2019_12_19_14_57_47_271712.csv')
Data2021 <- read_xlsx(
  '2021Fall_TreePhenologyData_All_wPhotoLinks.xlsx')
```

## Merge data

### Merge 2017-2019

```{r}
Data171819 <- Data2017 %>% 
  bind_rows(Data2018) %>% 
  bind_rows(Data2019)
```

### Species columns 

```{r, message = FALSE}
# Remove \n from the SPECIES column
Data2021 %<>% 
  mutate(SPECIES = str_remove(SPECIES, '\n'))

# Take a look at the two species columns
Data2021$SPECIES %>% table
Data171819$species %>% table

# Create dataframe with both styles of species names
SpeciesLookUp <- tibble(
  SPECIES = unique(Data2021$SPECIES) %>% sort,
  species = c(unique(Data171819$species), 'OSVI') %>% sort)

# Add the species names to both spreadsheets
Data171819 %<>% 
  left_join(SpeciesLookUp)
Data2021 %<>% 
  left_join(SpeciesLookUp)

```

### Leaf color columns

Will probably want to round to the nearest 5% since most people did so.

```{r, message = FALSE}
Data171819$color %>% table
# ggplot(Data171819, aes(x = color)) +
#   geom_bar() + tbw + th
Data2021$`LEAF COLOR` %>% table
# Note: bars barely show up due to 0.01 width
# ggplot(Data2021, aes(x = `LEAF COLOR`)) +
#   geom_bar() + tbw + th
```

### Leaf fall columns

I'm not sure how students found that a negative percent of leaves were falling. We can remove those three rows since we have replication and don't know what to do with them. There is also a lot of precision, 0.0001 recorded in 2021. We'll need to round those values.

```{r}
Data171819$fall %>% table
Data2021$`LEAF FALL` %>% table
```

### Accession number columns

140 trees are in both datasets. 36 trees are only in one dataset.

```{r}
# Most individual/ACCESSION values are in both dataframes
unique(Data171819$individual) %in% 
  unique(Data2021$ACCESSION) %>%
  table

unique(Data2021$ACCESSION) %in% 
  unique(Data171819$individual) %>%
  table

# Look at the values that are NOT in both data frames
unique(Data171819$individual)[!unique(Data171819$individual) %in% 
  unique(Data2021$ACCESSION)]

unique(Data2021$ACCESSION)[!unique(Data2021$ACCESSION) %in% 
  unique(Data171819$individual)]
```

### Photo links

```{r}
Data171819$photo %>% head
Data2021$`PICTURE ADDRESS` %>% head


```



### Rename columns as needed

```{r}
# Rename columns
Data171819 %<>% 
  rename(DATE = `_submission_time`)

Data2021 %<>%
  rename(color = `LEAF COLOR`,
         fall = `LEAF FALL`,
         individual = ACCESSION,
         photo = `PICTURE ADDRESS`)
```

### Add lat, long, altitude to 2021 data

Tried to do this, but it resulted in almost 500,000 rows of data due to rows of data with multiple lat/long/alt datasets. The problem is not just the altitude data, as removing that column doesn't help much.

```{r, message = FALSE}
Data2021 %>% 
  left_join(Data171819[,c('individual', '_tree_gps_latitude',
                         '_tree_gps_longitude', 
                         '_tree_gps_altitude')]) %>% dim

Data2021 %>% 
  left_join(Data171819[,c('individual', '_tree_gps_latitude',
                         '_tree_gps_longitude')]) %>% dim

```


### Merge select columns

```{r}
# Currently including group_num and student_id in case we want to use them as random effects. We don't have them for all years though
AllData <- Data171819[,c('SPECIES', 'species', 'individual',
                         'color', 'fall', 'DATE',
                         "_tree_gps_latitude",
                         '_tree_gps_longitude',
                         '_tree_gps_altitude',
                         'group_num', 'student_id',
                         'photo')] %>% 
  bind_rows(Data2021[,c('SPECIES', 'species', 'individual',
                        'color', 'fall', 'DATE',
                        # "_tree_gps_latitude",
                        # '_tree_gps_longitude',
                        # '_tree_gps_altitude',
                        'photo')])

# Rename lat/long/altitude for ease
AllData %<>% rename(Latitude = '_tree_gps_latitude',
                    Longitude = '_tree_gps_longitude',
                    Altitude = '_tree_gps_altitude')

```


## Remove questionable data

```{r}
# Negative % leaf fall
AllData %>% 
  filter(color < 0 |
         fall < 0)
# Remove negatives
AllData %<>% 
  filter(color >= 0,
         fall >= 0)
```

## Round leaf color and leaf fall data

```{r}
AllData %<>% 
  mutate(ColorR1 = round(color, digits = 0),
         ColorR5 = round(color / 5, digits = 0) * 5,
         ColorR10 = round(color, digits = -1),
         FallR1 = round(fall, digits = 0),
         FallR5 = round(fall / 5, digits = 0) * 5,
         FallR10 = round(fall, digits = -1))
AllData %>% 
  select(color, ColorR1, ColorR5, ColorR10, 
         fall, FallR1, FallR5, FallR10) %>% 
  unique %>% 
  head(20)
```

### Visualize

People seemed to naturally round to 5% in many cases, and somewhat to 10% as well. May want to use these smoothed values, though might not be necessary.

```{r}
ForPlot <- AllData %>% 
  pivot_longer(cols = c(ColorR1, ColorR5, ColorR10,
                        FallR1, FallR5, FallR10),
               names_to = 'Metric', values_to = 'Values')
ggplot(ForPlot %>% filter(Metric %in% c('ColorR1', 'FallR1')), 
       aes(x = Values)) +
  geom_bar() + 
  tbw + th + 
  facet_wrap(~ Metric)

ggplot(ForPlot %>% filter(Metric %in% c('ColorR5', 'FallR5')), 
       aes(x = Values)) +
  geom_bar() + 
  tbw + th + 
  facet_wrap(~ Metric)

ggplot(ForPlot %>% filter(Metric %in% c('ColorR10', 'FallR10')), 
       aes(x = Values)) +
  geom_bar() + 
  tbw + th + 
  facet_wrap(~ Metric)

```

## Date information

Extract date (YMD). Columns for year, month, day, week

```{r}
AllData %<>% 
  mutate(Year = year(DATE),
         Month = month(DATE),
         Day = day(DATE),
         Week = week(DATE))

AllData %>% 
  select(DATE, Year, Month, Day, Week) %>% 
  unique %>% 
  tail
```

## Unique observer

```{r}
# We have group_num and student_id for some sets of data. Not sure it'll be useful since we don't have them for everything. If needed though, here's a unique group_num/student_id/year variable
AllData %<>% 
  mutate(UniqueGroup = paste(Year, group_num, sep = "_"),
         UniqueStudent = paste(Year, group_num, student_id, 
                               sep = "_"))
```


## Export cleaned data

```{r}
# Commented out since it's already run.
# Uncomment if needed to re-export.
# write.csv(AllData, file = 'CleanedPhenologyData2017to2021.csv',
#           row.names = FALSE)

```

